---
title: "🧬 Spatial Transcriptomics Analysis"
format: html
---

# Welcome to the NGS-CN Summer School on Spatial Transcriptomics Analysis!

Spatial transcriptomics allows us to measure gene expression across tissue sections while preserving spatial information, helping us understand not just **what** genes are active, but **where** they are active.

In this practical session, we will explore how to analyze spatial transcriptomics data using **R** and **RStudio**. Whether you are new to spatial data or already familiar with single-cell workflows, this guide will walk you through the essential steps.

Feel free to use your personal workbook to make notes, track your progress, and revisit it after this workshop.

## Setting up your R environment 

Before diving into the data, we need to load a few important R packages. These libraries will provide the tools we need for analysis and visualization.

```{r}
#| label: setup
#| message: false

library(Seurat)     # Main package for single-cell and spatial transcriptomics analysis
library(patchwork)  # Helps combine multiple ggplot2 plots into one figure
library(tidyverse)  # Essential collection of R packages, including ggplot2, dplyr, magrittr
library(grid)       # For other plots
library(viridis)    # Our continuous colour palette

# Colours
cal_pal50 = c("#Fa1a8e", "#009B7D", "#ff9933", "#7083a4", "#ffce45", "#015e05", "#fedf25", 
"#d2b48c", "#bb55e1", "#6ec5ab", "#5d57af", "#143341", "#761445", "#d65b5a", 
"#94043a", "#e7a6cd", "#204519", "#87afd1", "#9b9a9d", "#f95b70", "#83c874", 
"#808080", "#452b5b", "#ecb100", "#f46124", "#525252", "#4c84a3", "#00bfff", 
"#01b4c6", "#174d79", "#a6a0f2", "#76facc", "#8491be", "#a32a2f", "#1c8859", 
"#2cc012", "#35782b", "#9c6755", "#3b3960", "#eeb0a1", "#3e1e13", "#0064c3", 
"#d81e4a", "#74646c", "#f675da", "#ffce45", "#ec7014", "#e50000", "#000000", 
"#a4527c", "#041859")   

# ggplot2 defaults
options(ggplot2.discrete.colour=cal_pal50)
options(ggplot2.discrete.fill=cal_pal50)
options(ggplot2.continuous.colour="viridis")
options(ggplot2.continuous.fill="viridis")
options(ggplot2.binned.colour="viridis")
options(ggplot2.binned.fill="viridis")
```

# Our dataset

In this section, we will introduce the dataset we are using for this training session.

## Dataset overview

We use a **mouse brain dataset** from [10x Genomics](https://www.10xgenomics.com/):   

* Source: Downloaded from [here](https://www.10xgenomics.com/datasets/visium-hd-cytassist-gene-expression-libraries-of-mouse-brain-he)   
* Sample: Male C57BL/6 mouse, 8 weeks old   
* Tissue: Brain tissue block (Formalin-Fixed Paraffin-Embedded, or FFPE)   
* Platform: Visium HD Spatial Gene Expression kit   
* Sequencing: Illumina NovaSeq 6000, approximately 300 million reads   
* Processing: Space Ranger was already run to process the raw data   

The dataset is located in the `datasets/visiumhd_mouse_brain` directory within your project folder. If you need to download the dataset again, you can use the `datasets/download.R` script.

## Dataset files

The dataset includes several files and folders:

* `web_summary.html`: Quality control (QC) report in HTML format
* `metrics_summary.csv`: QC report in CSV format
* `cloupe_008um.cloupe`: Loupe file for visualizing the data in the Loupe Browser
* `molecule_info.h5`: H5 file with information about reads and molecules in the dataset
* `feature_slice.h5`: H5 file for high-performance data access of 2 µm bins
* `spatial`: Folder containing the tissue images
* `binned_outputs` Folder containing results organized by different bin sizes
   * `square_002um/`: Binned data with 2 µm resolution
   * `square_008um/`: Binned data with 8 µm resolution
   * `square_0016um/`: Binned data with 16 µm resolution

The folder `binned_outputs` is relevant for our analysis. Space Ranger defaults to grouping the data into square bins of 2 µm, 8 µm, and 16 µm, though these bin sizes can be modified in the Space Ranger settings.

## Focusing on `square_008um`

Let us focus on the `square_008um/` folder, which contains the binned results at 8 µm resolution. This folder includes:

* `analysis/`: Results from Space Ranger analysis
* `cloupe.cloupe`: Loupe file for visualizing the 8 µm binned data in Loupe Browser
* `raw_feature_bc_matrix/`: Raw counts for all bins (tissue or no tissue)
* `raw_feature_bc_matrix.h5`: As above, organized in H5 format
* `filtered_feature_bc_matrix/`: Counts for all bins located under tissue, per bin and gene
* `filtered_feature_bc_matrix.h5`: As above, organized in H5 format
* `spatial/`:  Folder with images and information about the spatial locations of the bins on the tissue

The key files for our analysis are located in `filtered_feature_bc_matrix`. 

## Counts data

Let us take a closer look at the `filtered_feature_bc_matrix/` directory, which includes three important files:

* `barcodes.tsv.gz`: Contains the names (or identifiers) for each of the bins in the dataset
* `features.tsv.gz` Contains essential information about the genes, including their Ensembl IDs, symbols and gene types
* `matrix.mtx.gz` Contains the actual counts data
   * The first row contains the total number of genes, bins and counts.
   * All other rows contain the gene index (row in `features.tsv.gz`), bin index (row in `barcodes.tsv.gz`) and the count value.
   * Only non-zero counts are stored.

```{r}
# barcodes.tsv.gz
read.delim("datasets/visiumhd_mouse_brain/binned_outputs/square_016um/filtered_feature_bc_matrix/barcodes.tsv.gz", header = FALSE, nrows = 5)

# features.tsv.gz
read.delim("datasets/visiumhd_mouse_brain/binned_outputs/square_016um/filtered_feature_bc_matrix/features.tsv.gz", header = FALSE, nrows = 5)

# matrix.mtx.gz
read.delim("datasets/visiumhd_mouse_brain/binned_outputs/square_016um/filtered_feature_bc_matrix/matrix.mtx.gz", skip = 2, nrows = 5, header = FALSE, sep = " ")
```

::: {.callout-tip}
## Example
This dataset has 19,059 genes, 98,917 bins and 68,259,814 counts.  
Row 2: Gene #**17** (Vxn) has in bin #**1** (s_016um_00052_00082-1) a count of **3**.
:::

# The Seurat object

The R package [Seurat](https://satijalab.org/seurat/) is a widely used tool for single-cell and spatial transcriptomics analysis. It allows researchers to work with complex biological data and perform tasks like dimensionality reduction, clustering, and visualization. 

Another popular option is the [scanpy](https://squidpy.readthedocs.io/en/stable/) package in Python. 

## Importing the data into Seurat

The first step in analyzing your data is to import it into Seurat. Here, we load data for both 8µm and 16µm bins into a Seurat object: 

```{r}
sc = Load10X_Spatial("datasets/visiumhd_mouse_brain", bin.size=c(8, 16), assay="RNA")
sc$orig.ident = "brain"
Idents(sc) = "brain"
```

Now, let us take a look at the Seurat object to understand its structure:

```{r}
sc
```

::: {.callout-note}
## Questions
How many assays are in the dataset?  
How many images are in the dataset?
Why always two?
:::

## What is stored in the Seurat object?

The Seurat object organizes data into different components. These components facilitate spatial transcriptomics analysis and are crucial for data interpretation. The key components include:

- **Bins and genes**: These represent the spatial units (bins) and the genes that are being measured
- **Assay data**: Contains the gene expression data, chromatin accessibility, surface proteins, etc.
- **Metadata for bins**: Information about each bin (e.g., sample type, condition)
- **Metadata for genes**: Includes gene IDs, names, types, and descriptions
- **Dimensionality reductions**: Methods like PCA, UMAP, and tSNE that are used to reduce the complexity of the data
- **Neighbors and graphs**: Information about the relatedness between bins
- **Images**: Spatial coordinates and image data for visualizing the tissue

## Bins and genes

We can retrieve the total number of genes and bins stored in the Seurat object like this:

```{r}
dim(sc)
```

We can also list the bins and genes contained in the Seurat object:

```{r}
Cells(sc) %>% head(5)
Features(sc) %>% head(5)
```

::: {.callout-tip}
## Nomenclature
Note that sometimes the term **bin** is also referred to as barcode, cell, nuclei.   
The term **gene** is also referred to as feature, protein, peak.   
We will stick to bins and genes for this workshop. 
:::

Furthermore, bins have a default identity which is used for plotting. It can be accessed with the `Idents` function:

```{r}
Idents(sc) %>% head(5)
```

## Assay data

Assays store the actual data, for example gene expression counts. We can list the available assays in the Seurat object:

```{r}
Assays(sc)
```

To access a specific assay, we use the `GetAssay` function or the `[[]]` syntax:

```{r}
GetAssay(sc, assay="RNA.008um")
sc[["RNA.008um"]]
```

We can also check the number of genes and bins in a specific assay like this:

```{r}
assay = GetAssay(sc, assay="RNA.008um")
dim(assay)
```

::: {.callout-note}
## Questions
How many bins are in the assay?   
How many genes are in the assay?
:::

The assay object stores data in different **layers**:  

* `counts`: Raw counts for each gene in each bin  
* `data`: Normalized counts for better comparability across bins  
* `scale.data`: Scaled data for better comparability across genes  

We use the `GetAssayData` function to access the actual data:  

```{r}
#| message: true

counts = GetAssayData(sc, assay="RNA.008um", layer="counts")
counts[1:10, 1:10]
```

::: {.callout-note}
## Questions
What do you find special about this matrix?
:::

The Seurat object has a default/active assay that is used for most functions. To check the active assay, we can use:

```{r}
DefaultAssay(sc)
```

To change the active assay to another one, for example `"RNA.016um"`, use:

```{r}
DefaultAssay(sc) = "RNA.016um"
DefaultAssay(sc)
```

For this workshop, we will focus on the assay `"RNA.008um"` for our analysis:

```{r}
DefaultAssay(sc) = "RNA.008um"
DefaultAssay(sc)
```

::: {.callout-tip}
## Default assay
Most Seurat functions will automatically use the default assay, but you can always specify a different assay with the `assay=` argument.
:::

## Bin metadata

Bin metadata store additional information about each bin, such as the sample or condition. You can access this metadata using the `[[]]` syntax:

```{r}
sc[[]] %>% head(5)
```

For example, the `orig.ident` column stores the original identity of the bin, and the `nCount_RNA.008um` and `nFeature_RNA.008um` columns store the number of counts and genes for the 8µm bins.

You can access individual metadata columns like this:

```{r}
sc$nCount_RNA.008um %>% head(3)
```

## Gene metadata

Gene metadata can be accessed using the `[[]]` syntax on the Seurat assay object:

```{r}
#| message: true

assay <- GetAssay(sc, assay = "RNA.008um")
assay[[]] %>% head(5)
```

In this example, no gene metadata is available for the assay. If available, you can access individual columns using the `$` syntax.

## Dimensionality reductions

Dimensionality reductions (e.g., PCA, UMAP, tSNE) are essential for visualizing high-dimensional data in a lower-dimensional space. You can list available reductions with the `Reductions` function:

```{r}
Reductions(sc)
```

Currently, there are no reductions in the Seurat object. If they were present, you could access them using the `GetReduction` function or the `[[]]` syntax.

## Neighbors and graphs

To identify how the bins are related (e.g., for clustering), Seurat computes neighbors and graphs. You can list available neighbors and graphs using:

```{r}
Neighbors(sc)
Graphs(sc)
```

Currently, there are no computed neighbors or graphs. If available, you could access them with the `GetNeighbors` / `GetGraph` functions or the `[[]]` syntax.

## Images

For spatial data, images are stored in the Seurat object to visualize tissue sections and spatial coordinates. To list available images, use:

```{r}
Images(sc)
```

Each assay and bin size should have a corresponding image. You can access a specific image using the `[[]]` syntax:

```{r}
sc[["slice1.008um"]]
```

When we say image, what we mean are the **coordinates of an image**. We can also visualize the underlying **raw image** like: 

When we refer to the image in spatial transcriptomics, we usually mean the spatial coordinates associated with each bin. These coordinates tell us where each bin is located within the tissue section but not what the tissue looks like visually.

However, Seurat also allows us to view the underlying histological image that was used to derive these spatial coordinates. This can help us visually inspect the tissue morphology in parallel with gene expression patterns.

Here is how we can visualize the raw tissue image grid along with the spatial:
H&E Staining
dark = 


```{r}
#| fig-width: 5

grid::grid.newpage()
grid::grid.draw(GetImage(sc[["slice1.008um"]], method="grob"))
```

## Plotting

Now that we have the data in Seurat, we can start visualizing gene expression using plots.

### Violin plot for gene expression

We can create a **violin plot** to show the expression of specific genes (e.g., "Nrgn", "Igf2", "Hap1") in the RNA.008um assay:

```{r}
VlnPlot(sc, features = c("Nrgn", "Igf2", "Hap1"), layer="counts", pt.size = 0) + NoLegend()
```

Since the counts are raw and not normalized, interpreting them directly can be challenging.

### Spatial plot for gene expression

We can also visualize gene expression in a spatial context using a **spatial feature plot**:

```{r}
SpatialFeaturePlot(sc, features = c("Nrgn", "Igf2", "Hap1"), slot = "counts")
```

Note that `SpatialFeaturePlot` is for numeric values but there is also `SpatialDimPlot` which is for categorial values. 

### Comparison Between 8µm and 16µm Bins

Finally, we can plot the gene expression in the **RNA.016um assays**. This allows us to visualize how the data changes when comparing different bin sizes.

```{r}
DefaultAssay(sc) <- "RNA.016um"
p1 <- VlnPlot(sc, features = c("Nrgn"), layer="counts", pt.size = 0) + NoLegend()
p2 <- SpatialFeaturePlot(sc, features = c("Nrgn"), slot = "counts")
DefaultAssay(sc) <- "RNA.008um"
p1 | p2
```

# Quality control

## Why quality control?

In spatial transcriptomics (and any sequencing-based assay), not all bins are equally informative. Quality Control (QC) is used to identify and remove low-quality bins from the dataset before further analysis. 

**Low-quality bins** might be the result of:  

* Problems during reverse transcription, PCR amplification or library preparation  
* Cell damage during the experiment, e.g. when the tissue already contains a lot of dead or dying cells  
* Very few transcripts overall, e.g. in empty regions  
* Unusually high numbers of transcripts, possibly doublets or noise  

**Why does this matter?**  
Low-quality bins can distort the results. For instance, they might cluster separately and be falsely interpreted as novel biological states or cell types. Performing QC early helps ensure the reliability of downstream steps like normalization, clustering, and cell type annotation.

## QC metrics at a glance

We will use three common metrics to flag low-quality bins:

- **Number of counts (UMIs):** Measures total expression in a bin. Too low = empty; too high = potential artifacts.
- **Number of detected genes:** Captures the diversity of expression in a bin. Low gene counts might mean poor capture.
- **Mitochondrial gene content:** High levels often indicate stressed or dying cells.

These metrics will be assessed both statistically and spatially to get an idea of the sample quality and to identify thresholds for filtering.

## Number of Counts per Bin

The total number of UMIs (counts) per bin gives a sense of overall transcript abundance.

```{r}
p1 = VlnPlot(sc, features="nCount_RNA.008um", pt.size=0) + NoLegend()
p2 = SpatialFeaturePlot(sc, features="nCount_RNA.008um")
p1 | p2

summary(sc$nCount_RNA.008um)
```

Look for bins with very few or very many counts. Bins with <30 counts or >2000 counts are suspicious and flagged for removal. 

::: {.callout-tip}
## Bins flagged here should appear randomly across the tissue
If they form a distinct pattern, reconsider the threshold — you might be removing biologically relevant data.
:::

```{r}
filtered_bins_counts = WhichCells(sc, expression=nCount_RNA.008um < 30 | nCount_RNA.008um > 2000)
SpatialDimPlot(sc, cells.highlight=filtered_bins_counts, cols.highlight=c("#FFFF00", "grey50")) + NoLegend()
```

::: {.callout-note}
## Find your optimal threshold
Play around with the thresholds in R. Which threshold would you find best?
:::

## Number of Genes per Bin

This tells us how many genes are detected per bin — essentially a measure of complexity.

```{r}
p1 <- VlnPlot(sc, features="nFeature_RNA.008um", pt.size=0) + NoLegend()
p2 <- SpatialFeaturePlot(sc, features="nFeature_RNA.008um")
p1 | p2

summary(sc$nFeature_RNA.008um)
```

```{r}
filtered_bins_genes = WhichCells(sc, expression = nFeature_RNA.008um < 30 | nFeature_RNA.008um > 1000)

SpatialDimPlot(sc, cells.highlight=filtered_bins_genes, cols.highlight=c("#FFFF00", "grey50")) + NoLegend()
```

## Mitochondrial content

Mitochondrial RNA (mtRNA) often increases when cells are stressed or dying. A high percentage is a red flag.

We identify mitochondrial genes using a prefix (^mt-) and calculate their percentage per bin:

```{r}
sc = PercentageFeatureSet(sc, pattern="^mt-", col.name="pMito_RNA.008um") 

p1 = VlnPlot(sc, features="pMito_RNA.008um", pt.size=0) + NoLegend()
p2 = SpatialFeaturePlot(sc, features="pMito_RNA.008um") + NoLegend()
p1 | p2
```

```{r}
filtered_bins_mito = WhichCells(sc,
                                  expression = pMito_RNA.008um > 30)
SpatialDimPlot(sc, cells.highlight=filtered_bins_mito, cols.highlight=c("#FFFF00", "grey50")) + NoLegend()
```

## Filtering of bins

Once you have reviewed the metrics and decided on thresholds, remove poor-quality bins using subset():

Before filtering:

```{r}
dim(sc[["RNA.008um"]])
```

Apply the filters:

```{r}
sc = subset(sc, nCount_RNA.008um < 30 | nCount_RNA.008um > 2000, invert=TRUE)
sc = subset(sc, nFeature_RNA.008um < 30 | nFeature_RNA.008um > 1000, invert=TRUE)
sc = subset(sc, pMito_RNA.008um > 30, invert=TRUE)
sc[["RNA.008um"]]
```

## Filtering of genes

Genes expressed in only a few bins can add noise. First, we count how many bins each gene is detected in:

```{r}
counts = GetAssayData(sc, layer="counts")
genes_expressed = rowSums(counts > 0)
head(genes_expressed, 5)
summary(genes_expressed)
```

Let us quickly summarize these numbers. Most of the genes are expressed in at least 30 bins, on average in 4063 bins and maximum in 283994 bins.

```{r}
summary(genes_expressed)
```

Filter out genes expressed in ≤10 bins:

```{r}
genes_expressed = genes_expressed[genes_expressed > 10]
length(genes_expressed)
```

Then subset the Seurat object:

```{r}
sc[["RNA.008um"]]
sc[["RNA.008um"]] = subset(sc[["RNA.008um"]], features=names(genes_expressed))
sc[["RNA.008um"]]
```

::: {.callout-tip}
## Best practices
* Avoid "one-size-fits-all" thresholds; adjust based on sample type, bin size, and sequencing depth
* Make sure to check the spatial distribution of filtered bins to avoid removing spatial patterns
* At the beginning, run a full analysis (until clustering) without filtering 
  * Low-quality bins often cluster together
  * Plot QC metrics per cluster and identify clusters without or with ambiguous cell type
  * Then adjust filtering criteria accordingly and repeat
* Advanced: use tools that include neighboring bins (e.g [SpotSweeper](https://github.com/MicTott/SpotSweeper))
:::

::: {.callout-note}
## Questions  
What might cause unusually high count bins?  
Why might mitochondrial content be misleading in some tissues (e.g., muscle or brain)?  
What are the trade-offs of stricter vs. more lenient filtering?  
:::

```{r}
knitr::knit_exit()
```

# Preprocessing

After removing low-quality bins, we now prepare the dataset for downstream analysis. This **preprocessing pipeline** includes:

1. Normalization of raw counts
2. Identification of highly variable genes (HVGs)
3. Centering and scaling of those HVGs
4. Optional sketching of the dataset for speed (HVGs)
5. Dimensionality reduction (PCA) based on the HGVs

Each step ensures that only informative, reliable data is passed into clustering and annotation workflows.

## Normalization

Raw counts vary widely across bins due to differences in sequencing depth (library size). To make counts **comparable across bins**, we perform normalization.

Note that normalization methods for spatial data are still being developed and evaluated. For this workshop, we apply log-normalization. 

**How does it work?**  

* Per cell, the count for each gene is divided by the total count for that cell, and then multiplied by 10,000  
* These "counts per 10,000" are then log-transformed using the natural logarithm (with a pseudocount of 1)  

This is implemented in Seurat with `NormalizeData()`:

```{r}
sc = NormalizeData(sc, normalization.method="LogNormalize", scale.factor=10000)
sc[["RNA.008um"]]
```

This creates a new layer in the assay object called `data` that contains the normalized counts. You can inspect the normalized values like so:

```{r}
data = GetAssayData(sc, layer="data")
data[1:10, 1:10]
```

## Identifying Highly Variable Genes (HVGs)

Not all genes are equally informative. Experience shows that 1,000-3,000 genes with the highest bin-to-bin variation are often sufficient to describe the global structure of a dataset. 

Highly variable genes are typically the genes that show strong variation across bins and are more likely to distinguish biological states. Housekeeping genes with similar expression profiles across all bins, or genes with minor expression differences, might add random noise and mask relevant changes during downstream analysis. 

Here, we determine the top 2000 HGVs: 

```{r}
sc = FindVariableFeatures(sc, selection.method="vst", nfeatures=2000)
```

Get the top 10 HVGs and plot them:

```{r}
top10 = VariableFeatures(sc) %>% head(10)
top10

p = VariableFeaturePlot(sc, selection.method="vst")
p = LabelPoints(plot=p, points=top10, repel=TRUE)
p
```

## Data scaling

Gene expression ranges vary; some genes have inherently higher values. Centering and scaling ensures each gene contributes equally during the analysis. By centering (mean expression of 0) and scaling (standard deviation of 1), all genes will be on the same scale which makes them comparable. 

We use the `ScaleData` function to center and scale HGVs:

```{r}
sc = ScaleData(sc, features=VariableFeatures(sc))
GetAssay(sc, assay="RNA.008um")
```

This creates a new layer in the assay object called `scale.data` that contains the scaled counts:

```{r}
scale_data = GetAssayData(sc, layer="scale.data")
scale_data[1:5, 1:5]
```

## Creating a sketch assay

Large datasets can slow down clustering and UMAP. A sketch assay creates a reduced representation while preserving rare cell populations.

Instead of random sampling, Seurat uses a leverage score to keep influential or rare bins. We sketch 50,000 bins:

```{r}
# we select 50,0000 cells and create a new 'sketch' assay
sc = SketchData(sc, ncells=50000, method="LeverageScore", sketched.assay="sketch", features=VariableFeatures(sc))
sc
```

When we print the Seurat object, we see that the default is now the sketch assay. We can also check that it is indeed smaller:

```{r}
GetAssay(sc, assay="sketch")
```

Repeat HVG selection and scaling for the sketch:

```{r}
sc = FindVariableFeatures(sc, selection.method="vst", nfeatures=2000)
sc = ScaleData(sc, features=VariableFeatures(sc))
GetAssay(sc, assay="sketch")
```

::: {.callout-tip}
## Tip
Sketching is ideal for exploration on (very) large datasets. If your computing machine is super powerful or your dataset reasonably small, you might not need it. 
:::

## Dimensionality reduction

With ~20,000 genes, the data is too high-dimensional for visualization or clustering. At this point of the analysis, we have already reduced the dimensionality of the dataset to 2,000 HGVs. The biological manifold however can be described by far fewer dimensions than the number of (variable) genes, since expression profiles of different genes are correlated if they are involved in the same biological process. 

We use Principal Component Analysis (PCA) to reduce the dimensionality. PCA finds new axes (PCs) that capture the most variance in the dataset, and projects the data onto these axes. The resulting meta-variables are called principal components (PCs).

We use the `RunPCA` function to calculate the top 50 principle components (PCs) for the data. This creates a new dimensionality reduction object called `pca.sketch`:

```{r}
sc = RunPCA(sc, reduction.name="pca.sketch", features=VariableFeatures(sc), npcs=50)
Reductions(sc, slot="pca.sketch")
```

Visualize first two PCs:

```{r}
DimPlot(sc, reduction="pca.sketch", dims=c(1, 2))
```

In PCA, genes with the largest variation between bins will have the biggest influence on the PCs. The first PC captures the greatest variance across the dataset. The next PC captures the greatest remaining amount of variance, and so on. This way, the top PCs are likely to represent the biological signal where multiple genes are affected by the same biological processes in a coordinated way. In contrast, random technical or biological noise that affects each gene independently are contained in later PCs. 

Downstream analysis can be restricted to the top PCs. We need to decide how many PCs we want to use for our analysis. The following elbow plot shows PCs ranked based on the variance they explain. 

Look for the "elbow" — a point where adding more PCs gives diminishing returns:

```{r}
ElbowPlot(sc, reduction="pca.sketch", ndims=50)
```

You can also check which genes contribute most to each PC:

```{r}
VizDimLoadings(sc, reduction="pca.sketch", dims=1:4, nfeatures=10, balanced=TRUE)
```

::: {.callout-note}
## Question
How many PCs would you choose for further analysis? 
:::

# Clustering and visualization

- a
- b


## Clustering of the sketched dataset

For clustering, Seurat applies a graph-based approach. In short, Seurat first identifies for each bin its k nearest neighbors based on euclidean distance in PCA space. Using this neighborhood information, it then builds a nearest-neighbors graph (NN) and an improved version called shared nearest neighbor (SNN) graph.

We construct these graphs with the `FindNeighbors` function. We use the first 30 PCs as input and set the number of nearest neighbors to 20:

```{r}
sc = FindNeighbors(sc, reduction="pca.sketch", dims=1:30, k.param=20)
Graphs(sc)
```

Seurat then splits the SNN graph into well-connected communities. It does this by iteratively grouping neighboring bins together while trying to optimize the modularity. Modularity is is a measure for connectivity in graphs and a high value indicates that the bins within a community are more connected to each other than to bins outside of the community. The final communities are called clusters.

We run the `FindClusters` function and use the Leiden algorithm instead of the default Louvain algorithm (`algorithm=4`). The resolution of the clustering can be controlled with the `resolution` parameter. Higher values lead to more clusters. We use `resolution=2` as starting point:

```{r}
sc = FindClusters(sc, resolution=2, algorithm=4)

# Save as extra metadata column
sc$seurat_clusters.sketch = sc$seurat_clusters
```

The clusters are saved as bin metadata column `seurat_clusters.sketch`. Let us summarise the clustering of the sketched dataset:

```{r}
table(sc$seurat_clusters.sketch)
```

Note that we can provide multiple resolution values to test different clustering resolutions.

## UMAP of the sketched dataset

Additionally, we run a dimensionality reduction method called Uniform Manifold Approximation and Projection (UMAP). UMAP is a non-linear method that embeds high-dimensional data into 2D while trying to preserve the local structure of the data. It is often used for visualization. However it should be avoided to draw biological conclusions from the UMAP only without actual clustering.

We run the `RunUMAP` function and use the first 30 PCs as input. This creates a new dimensionality reduction object called `umap.sketch`:

```{r}
sc = RunUMAP(sc, reduction="pca.sketch", reduction.name="umap.sketch", dims=1:30, return.model=TRUE)
```

Note that `umap.sketch` is now the default dimensionality reduction. We can plot the clustering on the UMAP as well as in spatial context:

```{r}
p1 = DimPlot(sc, reduction="umap.sketch", label=TRUE) + NoLegend()
p2 = SpatialDimPlot(sc, images="slice1.008um") + NoLegend() + scale_fill_manual(values=cal_pal50)
p1 | p2
```

Here we can nicely see that we only analysed a subsampled dataset.

## Analysing the full dataset

We now use the `ProjectData`function to project the cluster labels, PCA and UMAP, that we learned from the sketch assay, to the entire dataset:

```{r}
sc = ProjectData(sc,
                 sketched.assay="sketch", assay="RNA.008um", 
                 sketched.reduction="pca.sketch", full.reduction="pca.008um", 
                 umap.model="umap.sketch",
                 refdata=list(seurat_clusters.008um="seurat_clusters.sketch"),
                 dims=1:30)

# Fix the category levels for bin metadata column seurat_clusters.008um and update the bin identities
sc$seurat_clusters.008um = factor(sc$seurat_clusters.008um, levels=levels(sc$seurat_clusters.sketch))
Idents(sc) = "seurat_clusters.008um"

# Rename UMAP from full.umap.sketch to umap.008um
umap = sc[["full.umap.sketch"]]
Key(umap) = "umap.008"
sc[["umap.008um"]] = umap
sc[["full.umap.sketch"]] = NULL

# Set default assay back to RNA.008um
DefaultAssay(sc) = "RNA.008um"
```

The Seurat object now contains a full PCA (`pca.008um`), a full UMAP (`umap.008um`) and the full clustering (`seurat_clusters.008um`). Now, we plot the clustering of the full dataset on the UMAP as well as in spatial context:

```{r}
p1 = DimPlot(sc, reduction="umap.008um", label=TRUE) + NoLegend()
p2 = SpatialDimPlot(sc) + NoLegend() + scale_fill_manual(values=cal_pal50)
p1 | p2
```

When we have so many different clusters (of which some are spatially restricted and others not), it can be challenging to interpret their spatial location. Here it helps to highlight individual clusters:

```{r}
cluster_bins = CellsByIdentities(sc, idents = c(16, 8))
cluster_bins[["NA"]] = NULL

SpatialDimPlot(sc, cells.highlight=cluster_bins, facet.highlight=TRUE, cols.highlight=c("#FFFF00", "grey50")) + NoLegend()
```

## Marker genes

Once we have clustered the data, we can identify marker genes for each cluster. Marker genes are genes that are differentially expressed in a specific cluster compared to all other clusters.

## Plotting

We can explore the results with different plotting functions. We can plot gene expression (or other numeric values) per cluster as violin plot to get an idea of their distribution:

```{r}
VlnPlot(sc, features=c("Acta2", "Apoe"), pt.size=0, ncol=1) + NoLegend()
```

We can also plot them on top of our UMAP visualization:

```{r}
FeaturePlot(sc, features=c("Acta2", "Apoe"), pt.size=0)
```

And of course also in spatial context:

```{r}
SpatialFeaturePlot(sc, features=c("Acta2", "Apoe"), pt.size=0)
```

More comprehensive options are the dot plot and the heatmap. They allow us to visualize e.g. expression of multiple genes for multiple cluster in one plot. Here we plot the expression of three genes for clusters 1 to 10 as dot plot:

```{r}
DotPlot(sc, idents=1:10, 
        features=c("Hap1", "Nrgn", "Snap25", "Camk2n1", "Apoe", "Nap1l5", "Olfm1", "Penk", "Ptgds", "Pcp4"),
        scale=TRUE) + 
  scale_color_viridis() +
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

We can also plot the same data as heatmap. However, we can do this only for a subset of cells:

```{r}
set.seed(1)
cluster_bins = Cells(sc) %>% sample(10000)

DoHeatmap(sc, features=c("Hap1", "Nrgn", "Snap25", "Camk2n1", "Apoe", "Nap1l5", "Olfm1", "Penk", "Ptgds", "Pcp4"), 
          cells=cluster_bins,
          group.colors=cal_pal50) + 
  scale_fill_viridis() +
  guides(colour="none")
```

## Marker genes

## Subclustering

# Spatial domains



# Other useful resources

- [OSTA Visium HD Workflow](https://lmweber.org/OSTA/pages/seq-workflow-visium-hd.html)
- [Seurat Visium HD Analysis Vignette](https://satijalab.org/seurat/articles/visiumhd_analysis_vignette)


