---
title: "Our first workbook"
---

# Quarto

**Quarto** enables you to weave together content and executable code into a finished document. To learn more about *Quarto* see <https://quarto.org>.

## Run code in chunks

Chunks are the building blocks of a Quarto document. A chunk is a section of code that can be executed and its output included in the final document. Chunks are defined by three backticks followed by `{r}`, then the code, and finally three backticks again. For example:

```{r}
#| label: chunk_1

1 + 1
```

Run this chunk by clicking on the play symbol (top-right button) of the chunk. The output is shown below and can be anything: text, numbers, plots, ... . You can also run all chunks above this chunk clicking on the button next to the play symbol.

Chunks can have labels, which are useful for referencing them later. For example, the chunk above has the label `chunk_1`:

```{r}
#| label: chunk_2

plot(cars)
```


You can tune the behavior of the chunk by adding options. For example, you can add `#| echo: false` to hide the code in the output:

```{r}
#| label: chunk_3
#| echo: false

2 * 2
```

The most basic options can be set using the gear-shaped symbol in the top-right corner of the chunk. Other options can be found in the [knitr documentation](https://yihui.org/knitr/) or in the [quarto documentation](https://quarto.org/docs/guide/).

## Format your document content

Quarto documents are written in Markdown, a lightweight markup language that allows you to format text easily. You can create headings, lists, links, and more using simple syntax. For example, in this document:

- we use `#` for the title and `##` for the section titles
- we use `**` to print the word Quarto in **bold** and `*` to print it in *italic*
- we highlight words with single backticks `like this`
- we create links using the syntax `[link text](URL)`
- we use `-` to create a bullet list

A detail documentation of the Markdown syntax can be found in the [quarto documentation](https://quarto.org/docs/guide/).

## Create a final document

Once you are done, you can render a complete HTML document that contains all your documentation, code and results. Click the **Render** button at the top of the quarto file. Quarto also supports pdf and word output, lets you create presentations, websites, and more. You can find all the options in the [quarto documentation](https://quarto.org/docs/guide/).


